---
title: "Weather events most harmful to health and wealth"
author: "Tanveer Ansari"
date: "04/22/2015"
output: html_document
---

`r require(knitr);require(dplyr);require(ggplot2);require(reshape2);`
`r opts_chunk$set(cache=FALSE)`

```{r,results='hide',message=FALSE,warning=FALSE,include=FALSE}
library(dplyr);library(ggplot2);library(reshape2);
```
##Synopsis
Examine historical storm data for the United States to find out   
1) What types of storm events cause the most harm to population health?  
2) What types of storm events cause the most financial loss?  

Clean, summarize and analyze the data to find the most significant weather event types  
in both the above categories.

## DataProcessing  

###Strategy
* Read the data and perform any cleanup required  
* Categorize valid data by the type of event that occured  
* Group and summarize data by its event type to determine  
    a) The top event types that cause fatalities and injuries, ordered by number of fatalities.    
    b) The top event types that cause financial loss, ordered by total financial loss caused.  
    
*  Plot a bar chart of both the above

### Steps
Download file and read its data
```{r readAndLoad,cache=TRUE}
if(!file.exists("stormdata.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                destfile = "stormdata.csv.bz2")
  }

storm<-tbl_df(read.csv("stormdata.csv.bz2",stringsAsFactors=F,na.strings=c("NA,NAN","")))
```
===================================================================================

#### Analysis for top population health damage
Select only those events that caused either fatalities or injuries.  
Aggregate the data by eventtype, counting fatalities and injuries.  

```{r getinjuries,cache=TRUE}
injuries=subset(storm, FATALITIES>0 | INJURIES>0)
inj<-injuries%>% group_by(EVTYPE)%>%summarize(fatalities=sum(FATALITIES,na.rm = T),hurt=sum(INJURIES,na.rm=T))
```

Examine quantiles for fatalities so we can pick a reasonable cut-off point.

```{r}
quantile(inj$fatalities,prob=seq(0.95,1,by=0.01))
```

The 97th percentile of fatalities is ~400 and 98th is ~700 so choosing 500 as a cut-off should pick less than 3% of the categories.  
Select those event types that caused at least 500 fatalities or more and examine them sorted by most harmful first. 

```{r picklargeinj,cache=TRUE}
inj<-subset(inj,fatalities>500)
arrange(inj,desc(fatalities),desc(hurt))
```

Tornadoes are clearly the worst here.  



Order the eventtypes as a factor with levels in descending order of fatalities.  
This will make our plot ordered by fatalities, highest first.
```{r}
inj$EVTYPE=factor(inj$EVTYPE,levels=inj$EVTYPE[order(desc(inj$fatalities))])
```

Plot the data after melting, so we can use facets to split up fatalities vs injuries.  

```{r plotinjuries,fig.width=10}
injmelt<-melt(inj,id=c("EVTYPE"),measure.vars=c("fatalities","hurt"))

ggplot(injmelt,aes(x=EVTYPE,y=value,fill=factor(variable)))+geom_bar(position="dodge",stat="identity")+ggtitle(
  "Storm Events causing Top 6 Fatalities") +labs(y="Count") + labs(x="")
```
Plotting the fatalities and injuries per event it is evident that Tornadoes are the most dangerous event,
followed by excessive heat as a distant second.  

===================================================================================


#### Analysis for top economic consequence (property damange and crop damage)
To repeat a similar analysis for economic consequences, our data needs some cleaning up.

The dataset columns relevant to damage are split into xxDMG and xxDMGEXP where xx is PROP/CROP. 
The DMG column contains a number while EXP contains text. 

```{r getdamage,cache=TRUE}
stormdmg<-subset(storm,(PROPDMG>0 & !is.na(PROPDMGEXP))|(CROPDMG>0 & !is.na(CROPDMGEXP)))
## take only columns we are interested in
stormdmg<-select(stormdmg,EVTYPE,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```
Select the columns we are interested in for rows with non-zero damage numbers.  

Lets examine some of the suffixes in xxxEXP columns
```{r}
sort(unique(c(unique(stormdmg$CROPDMGEXP),unique(stormdmg$PROPDMGEXP))))
```
Some of the above suffixes appear to be invalid so our data needs cleaning up.  
From the description on the website 

>"Estimates should be rounded to three significant digits, followed by an alphabetical character   
>signifying the magnitude of the number, i.e., 1.55B for $1,550,000,000. Alphabetical characters  
>used to signify magnitude include "K" for thousands, "M" for millions, and "B" for billions"

However the values in the dataset also include symbols ike "?","2","-","+","m","M"   

For the purposes of this analysis both uppercase and lowercase k,m and b were considered 
to represent thousand, million and billion. All others were ignored by setting them to zero.

```{r filterDamageExp}
validDamage<-c("K","k","M","m","B","b")
hdmg<-filter(stormdmg,PROPDMGEXP %in% validDamage | CROPDMGEXP %in% validDamage)
```
Next we shall replace invalid suffixes with 0, "m"" with a million, "b"" with a billion and "k"" with a thousand,  
for both property and crop damageexp.

```{r replaceDamageExp, cache=TRUE}
#Replace NAs with zero
if(any(is.na(hdmg$PROPDMGEXP))) hdmg[which(is.na(hdmg$PROPDMGEXP)),]$PROPDMGEXP=0
#Replace K with a thousand and do the same for millions  and billions
hdmg$PROPDMGEXP<-gsub("[Kk]",as.character(10^3),hdmg$PROPDMGEXP)
hdmg$PROPDMGEXP<-gsub("[Mm]",as.character(10^6),hdmg$PROPDMGEXP)
hdmg$PROPDMGEXP<-gsub("[Bb]",as.character(10^9),hdmg$PROPDMGEXP)

# There are still some rows with bad values, which are 5 and 3 - replace this with 0
hdmg$PROPDMGEXP<-gsub("5",as.character(0),hdmg$PROPDMGEXP)
hdmg$PROPDMGEXP<-gsub("3",as.character(0),hdmg$PROPDMGEXP)


#Repeat above steps performed on property damage, for crop damage
if(any(is.na(hdmg$CROPDMGEXP))) hdmg[which(is.na(hdmg$CROPDMGEXP)),]$CROPDMGEXP=0
hdmg$CROPDMGEXP<-gsub("[Kk]",as.character(10^3),hdmg$CROPDMGEXP)
hdmg$CROPDMGEXP<-gsub("[Mm]",as.character(10^6),hdmg$CROPDMGEXP)
hdmg$CROPDMGEXP<-gsub("[Bb]",as.character(10^9),hdmg$CROPDMGEXP)

# There are still some rows with ? in CROPDMGEXP - replace this with 0
hdmg$CROPDMGEXP<-gsub("\\?",as.character(0),hdmg$CROPDMGEXP)
```



Now that the multipliers are fixed, we can get the actual number by multiplying CROPDMG with CROPDMGEXP and the same thing for PROPDMG. 
Lets also add add a totals column so we can sort events based on total damage caused, 

```{r calcDamangeTotal}
#Lets calculate the actual numbers in dollars and the total damage per observation
hdmg<-mutate(hdmg,prop=PROPDMG*as.numeric(PROPDMGEXP),crop=CROPDMG*as.numeric(CROPDMGEXP),total=crop+prop)
```


Aggregate incidents, summing property damage by event type. 

```{r arrangeDamage}
aggdmg<-hdmg%>%group_by(EVTYPE)%>%
  summarize(PropertyDamage=sum(prop),CropDamage=sum(crop),TotalDamage=sum(total))%>%arrange(desc(TotalDamage))
```

Take the top six most damaging event types and take a look at them
```{r getTopDamage}
#Take the top six
aggdmg<-aggdmg[1:6,]
aggdmg
```

Floods appear to be the worst.

Order the eventtypes as a factor with levels in descending order of total damage (existing order of data).  
This will make our plot ordered by total damage, highest first
```{r}
aggdmg$EVTYPE=factor(aggdmg$EVTYPE,levels=aggdmg$EVTYPE[order(desc(aggdmg$TotalDamage))])
```

Sorting is done so don't need the totals column anymore
```{r}
aggdmg$TotalDamage<-NULL
```

Plot the data after melting. The y scale is in millions for the sake of y-axis readability.

```{r plotdamage, fig.width=10}
dmg<-melt(data = aggdmg, id.vars = c("EVTYPE"), na.rm = T)
ggplot(dmg,aes(x=EVTYPE,y=value/10^6,fill=factor(variable)))+geom_bar(position="dodge",stat="identity")+
  ggtitle("Storm Events causing Top 6 Crop damage") + labs(y="Damage in Millions")+labs(x="")
```

===================================================================================


##Results

We find that Tornadoes cause the most public health damage and heat related  incidents are next.  
Flooding related events and hurricanes,typhoons and tornadoes caused a lot of economic damage.  
Tornadoes appear to be severely harmful being the Number 1 cause of fatalities and Number 3 for economic consequences.  
Flash floods were the only other event type that appeared in the top six event types for both categories.
