---
title: "TODO: Summarize data analysis"
author: "Tanveer Ansari"
date: "04/22/2015"
output: html_document
---
##Synopsis
  TODO: put synopsis here in AT MOST 10 sentences
## DataProcessing  
    TODO:Write description 
    TODO: show code (echo=TRUE, cache=TRUE)
    Note : Analysis MUST start from raw data
###Assumption about property damage and crop damage units
The columns are split into xxDMG and xxDMGEXP where xx is PROPor CROP. The DMG column contains a number  
while EXP contains text
From the description on the website  
>"Estimates should be rounded to three significant digits, followed by an alphabetical character   
>signifying the magnitude of the number, i.e., 1.55B for $1,550,000,000. Alphabetical characters  
>used to signify magnitude include "K" for thousands, "M" for millions, and "B" for billions"

However the values in the dataset also include symbols ike "?","2","-","+","m","M" 
For the purposes of this analysis both uppercase and lowercase k,m and b were considered 
to represent thousand, million and billion and all others ignored as NA


##Results
  TODO: Show Results here
  
#Checklist
1.  Check all TODOs
2.  PLOTS:  AT LEAST one plots. Figures may have more than one plot
3.  FIGURES: AT MOST  3 figures
4.  Publish on RPubs.com

========================================
using R Markdown see <http://rmarkdown.rstudio.com>.
```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
